---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  out.width = 500
)
```

# Theseus Plot: Visualizing Differences through Replacement in Rate Metrics

<!-- badges: start -->
<!-- badges: end -->

## 1. Overview

In data analysis, differences in metrics between groups often prompt questions about which attributes are responsible for the observed variation.
For instance, in A/B testing, you might want to assess whether gender differences contribute to the observed change in a key metric.
However, such analysis becomes challenging when the metrics are rates, as they cannot be aggregated across attributes in the same way volume metrics can.

## Installation

You can install the development version of Theseus like so:

``` r
remotes::install_github("hoxo-m/Theseus")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r}
library(dplyr)
library(nycflights13)

data <- flights |> 
  filter(!is.na(arr_delay)) |>
  mutate(on_time = arr_delay <= 0)  # arrived on time

data |> select(year, month, day, origin, dest, carrier, dep_delay, on_time) |> head()

data1 <- data |> filter(month == 9L)
data2 <- data |> filter(month == 12L)

data1 |> summarise(on_time_rate = mean(on_time)) |> pull(on_time_rate)
data2 |> summarise(on_time_rate = mean(on_time)) |> pull(on_time_rate)
```

```{r example}
library(Theseus)

ship <- create_ship(data1, data2, y = on_time, labels = c("2013-09", "2013-12"))

ship$table(origin)
```

```{r}
ship$plot(origin)
```

```{r}
ship$plot_flip(carrier)
```

```{r}
ship$plot_flip(dep_delay)
```
